A highly-available move operation for replicated trees

### 2019-06-17 paper/move-op.tex
paper: minor changes to description of get_parent
paper: more minor rewording as reading through...

Our contributions in this work are as follows:
\begin{itemize}
    \item We define a Conflict-free Replicated Data Type for trees that supports a \emph{move} operation, without requiring any coordination between replicas such as locking or consensus.
        As discussed in Section~\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
    \item We formalise our abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of our algorithm.
        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
    \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an efficient implementation within Isabelle/HOL and prove the equivalence of the two.  We thereafter extract a formally verified, executable Scala implementation from the Isabelle/HOL definitions, and perform tests on this implementation.
    \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
\end{itemize}

@@

We define only one kind of operation: \isa{Move t p m c} (Figure~\ref{fig:code}, lines 1--5).
A move operation is a record consisting of four fields: a timestamp \isa{t} of type \isa{'t}, a parent node \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node \isa{c} of type \isa{'n}.
Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be instantiated with arbitrary datatypes; our only requirement is that timestamps \isa{'t} are globally unique and have a total order.
For example, Lamport timestamps~\cite{Lamport:1978jq} meet these requirements.

@@

In order to correctly apply move operations, a replica needs to maintain not only the current state of the tree, but also an \emph{operation log}.
The log is a list of \isa{LogMove} records in descending timestamp order.
A \isa{LogMove t oldp p m c} record (lines 7--12) is similar to a record \isa{Move t p m c}; the only difference is that the \isa{LogMove} has an additional field \isa{oldp} of type \isa{('n {\isasymtimes} 'm) option}.
This \isa{option} type means the field can either take the value \isa{None} (equivalent to null), or a pair of an inode and a metadata field.

@@

In the first line of \isa{get\_parent}, the expression between \isa{::} and \textbf{where} is the type signature of the function, in this case:
\begin{quote}
\isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set {\isasymRightarrow} 'n {\isasymRightarrow} ('n {\isasymtimes} 'm) option}
\end{quote}
This signature denotes a function that takes two arguments: a tree \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set} and a node \isa{'n}.
It then returns a \isa{('n {\isasymtimes} 'm) option}.
\isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots''.

@@

Recall from Section~\ref{sec:move-cycle} that in order to prevent a cycle being introduced, the node being moved must not be an ancestor of the destination node.
To implement this in Figure~\ref{fig:code}, we first define the \isa{ancestor} relation in lines 22--24.
This relation is the transitive closure of the parent-child relation of a tree: if \isa{(p, m, c) {\isasymin} tree} then \isa{p} is an ancestor of \isa{c} (line 23); moreover, if \isa{(p, m, c) {\isasymin} tree} and \isa{a} is an ancestor of \isa{p}, then \isa{a} is also an ancestor of \isa{c} (line 24).
The \textbf{inductive} keyword indicates that this recursive definition is iterated until a fixed point is reached.

@@

The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
The expression \isa{'t::\{linorder\}} in the type signature indicates that the timestamps \isa{'t} belong to the type class \isa{linorder}, which means that timestamps can be compared with the \isa{<} operator defining a total order.
This comparison occurs on line 49.

>>>>

Our contributions in this work are as follows:
\begin{itemize}
    \item We define a Conflict-free Replicated Data Type for trees that supports a \emph{move} operation, without requiring any coordination between replicas such as locking or consensus.
        As discussed in Section~\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of the algorithm.
        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
    \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an efficient implementation within Isabelle/HOL and prove the equivalence of the two.  We thereafter extract a formally verified, executable Scala implementation from the Isabelle/HOL definitions, and perform tests on this implementation.
    \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
\end{itemize}

@@

We define only one kind of operation: \isa{Move t p m c} (Figure~\ref{fig:code}, lines 1--5).
A move operation is a record consisting of four fields: a timestamp \isa{t} of type \isa{'t}, a parent node \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node \isa{c} of type \isa{'n}.
Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be \emph{substituted} for arbitrary types; our only requirement is that timestamps \isa{'t} are globally unique and have an associated total order.
Concretely, Lamport timestamps~\cite{Lamport:1978jq} would meet these requirements.

@@

In order to correctly apply move operations, a replica needs to maintain not only the current state of the tree, but also an \emph{operation log}.
The log is a list of \isa{LogMove} records in descending timestamp order.
A \isa{LogMove t oldp p m c} record (lines 7--12) is similar to a record \isa{Move t p m c}; the only difference is that the \isa{LogMove} has an additional field \isa{oldp} of type \isa{('n {\isasymtimes} 'm) option}.
This \isa{option} type means the field can either take the value \isa{None} (intuitively similar to null), or a pair of an inode and a metadata field.

@@

In the first line of \isa{get\_parent}, the expression between \isa{::} and \textbf{where} is the type signature of the function, in this case:
\begin{quote}
\isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set {\isasymRightarrow} 'n {\isasymRightarrow} ('n {\isasymtimes} 'm) option}
\end{quote}
This signature denotes a function that takes two arguments: a tree \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set} and a node \isa{'n}.
It then returns a \isa{('n {\isasymtimes} 'm) option}.
\isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots'', whilst $\isa{THE}\ \isa{x}. \ldots$ means ``\emph{choose} the unique value \isa{x} such that\ldots''.

@@

Recall from Section~\ref{sec:move-cycle} that in order to prevent a cycle being introduced, the node being moved must not be an ancestor of the destination node.
To implement this in Figure~\ref{fig:code}, we first define the \isa{ancestor} relation in lines 22--24.
This relation is the transitive closure of the parent-child relation of a tree: if \isa{(p, m, c) {\isasymin} tree} then \isa{p} is an ancestor of \isa{c} (line 23); moreover, if \isa{(p, m, c) {\isasymin} tree} and \isa{a} is an ancestor of \isa{p}, then \isa{a} is also an ancestor of \isa{c} (line 24).
The \textbf{inductive} keyword indicates that this recursive definition is iterated until a fixed point (the least) is reached.

@@

The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and timestamps can therefore be compared with the \isa{<} operator defining a total order.
This comparison occurs on line 49.
